Sample-Efficient Reinforcement Learning \\ with Implicitly Quantized Representations

### 2024-02-02 paper/main.tex
reinforcement learning -> RL

How can high-dimensional and noisy observations be used efficiently in model-based RL without resorting to reconstruction or direct prediction of future frames? A common solution has been to attach an auxiliary loss to the RL objective and perform representation learning \cite{tomar2021learning, ni2024bridging}. A common approach for learning suitable representations is the use of self-predictive abstractions, where the model is trained to predict future latent states through an auxiliary loss \cite{subramanian2022approximate}. \citet{ye2021mastering} introduce a self-supervised consistency loss on the learned latent representations. Instead of relying on a reconstruction-based loss function, \citet{schwarzer2020data} propose a cosine similarity loss between predicted future latents and the true future latents and Q-learning in the learned latent space. \citet{ma2021contrastive} modify the classical world model learning objective by replacing the observation reconstruction loss with a contrastive learning term, which maximizes the mutual information between latent states and observations and helps combat the model focusing on learning task-irrelevant information. \citet{kipf2019contrastive} use contrastive learning to learn a structured world model. \citet{ghugare2022simplifying} argue that aligning the auxiliary objectives and RL objective can be difficult. The authors derive a bound directly on the overall RL objective, which is compatible with self-consistency. The contrastive objectives for model-based RL significantly improve sample efficiency, but they are not only limited to model-based reinforcement learning. \citet{laskinCURLContrastiveUnsupervised2020} propose replacing the transition function with data augmentations to get a contrastive target and performing model-free learning using the representation produced by the encoder trained with the contrastive target.

@@

\textbf{Reward prediction is not necessary for representation learning}
Unlike prior methods such as TCLR \cite{zhaoSimplifiedTemporalConsistency2023}, TD-MPC \cite{hansenTemporalDifferenceLearning2022}, and Dreamer-V2 \cite{hafnerMasteringAtariDiscrete2022}, our representation learning loss (\cref{eq:rep-loss}) does not include a term for learning to predict the reward or the value. Instead, we rely solely on the self-predictive temporal consistency loss. To analyze the impact of not including the reward prediction term, we perform an ablation study, where we compare our method to three variants of our method:
\begin{itemize}
	\item \ourrew, where we add a reward prediction head into our model architecture by re-defining the dynamics as $\hat{z}_{t+1}, \hat{r}_t = d_{\phi} (z_{t}, a_{t})$ (see \cref{eq:transition}), and include a reward prediction term (discounted MSE Loss) in the representation loss:
	\begin{align*}
		\mathcal{L}_\text{rew} = \sum_{h=0}^{H-1} \gamma^h \| \hat{r}_{t+h} - r_{t+h}\|_2^2,
	\end{align*}
	where $r_{t+h}$ is the ground-truth H-step reward and $\hat{r}_{t+h}$ the predicted H-step rewards.
	\item \ourcosrew, where in addition to the reward prediction head in \ourrew, we additionally convert the self-supervised MSE loss (see \cref{eq:rep-loss}) for the temporal consistency into a cosine similarity, similarly to TCRL:
	\begin{align*} \label{eq:rep-loss}
		\mathcal{L}_{\text{rep}}(\theta, \phi; \tau)
		&= - \sum_{h=0}^{H-1} \gamma^{h} \left ( \frac{\hat{z}_{t+h}}{\| \hat{z}_{t+h} \|_2} \right )^T \left ( \frac{\tilde{z}_{t+h}}{\| \tilde{z}_{t+h} \|_2} \right )
	\end{align*}
  where as before we have $\hat{z}_{t+h} = d_{\phi}(\hat{z}_{t}, a_{t})$ and $\tilde{z}_{t+h} = e_{\bar{\theta}}(s_{t+1})$.
	\item TCRL-ours, where in addition to \ourcosrew, we remove the FSQ normalization scheme to make the method equal to TCRL with some minor improvements to the method, such as using nstep-TD3 \cite{fujimotoAddressingFunctionApproximation2018} as the base algorithm instead of DDPG with Double Q and nstep returns, using the neural network architecture from \citet{hansen2023td}, and performing policy smoothing \cite{fujimotoAddressingFunctionApproximation2018}.
\end{itemize}
The results for this ablation study are shown in \cref{fig:reward-ablation}. 
The plots show that our method, \our, without a reward prediction term in the loss, has equal or superior performance to the comparison methods except in Fish Swim, where the unnormalized TCRL-ours outperforms all the methods with normalization due to being able to maintain the latent matrix rank in this specific environment (see \cref{fig:rank-comparison-z1}). Even then, \our performs equally to the normalized variants of our method with the reward prediction. Our results imply that learning to predict the reward is not necessary for learning a suitable latent representation. The upside of not including the reward prediction head is that it makes the representation task-agnostic, which we believe to be important for downstream applications such as learning in an incremental multi-task setting.

@@

\textbf{Reconstruction loss has a detrimental impact}
Learning to minimize the observation reconstruction error has been widely applied in model-based reinforcement learning \cite{sutton2018reinforcement, haRecurrentWorldModels2018, hafnerLearning2019}, and an observation decoder has been a component of many of the most successful RL algorithms to date \cite{hafner2023mastering}. However, recent work \citep{zhaoSimplifiedTemporalConsistency2023,hansenTemporalDifferenceLearning2022}
has shown that incorporating a reconstruction term into the representation loss can have a negative impact on performance, as learning to reconstruct the observations is inefficient due to the observations containing irrelevant details and visuals like shading that are uncontrollable by the agent and do not affect the tasks.

To provide a thorough analysis of \our, we include results where we add a reconstruction term to our representation loss \cref{eq:rep-loss}:
\begin{align*}
	\hat{o}_t &= g_\xi(z_t), \\
	\mathcal{L}_o &= \mathbb{E}_{o_t\sim\mathcal{D}} [\| \hat{o}_t - o_t \|_2^2],
\end{align*}
where $g_\xi$ is a learned observation decoder that takes the latent state as the input and outputs the reconstructed observation. The decoder $g_\xi$ is a standard MLP. We perform reconstruction at each time step in the horizon.
The results in \cref{fig:reconstruction-loss-ablation} show that in no environments does reconstruction aid learning, and in some environments, such as the difficult Dog Walk and Quadruped Run, including the reconstruction term has a significant detrimental effect on the performance.
This supports the observations of \citet{zhaoSimplifiedTemporalConsistency2023} and \citet{hansenTemporalDifferenceLearning2022} about the lack of need for a reconstruction target in continuous control tasks.
%Moreover, in the quadruped-run and dog-walk environments incorporating a reconstruction term has a detrimental impact on
%overall performance.

>>>>

How can high-dimensional and noisy observations be used efficiently in model-based RL without resorting to reconstruction or direct prediction of future frames? A common solution has been to attach an auxiliary loss to the RL objective and perform representation learning \cite{tomar2021learning, ni2024bridging}. A common approach for learning suitable representations is the use of self-predictive abstractions, where the model is trained to predict future latent states through an auxiliary loss \cite{subramanian2022approximate}. \citet{ye2021mastering} introduce a self-supervised consistency loss on the learned latent representations. Instead of relying on a reconstruction-based loss function, \citet{schwarzer2020data} propose a cosine similarity loss between predicted future latents and the true future latents and Q-learning in the learned latent space. \citet{ma2021contrastive} modify the classical world model learning objective by replacing the observation reconstruction loss with a contrastive learning term, which maximizes the mutual information between latent states and observations and helps combat the model focusing on learning task-irrelevant information. \citet{kipf2019contrastive} use contrastive learning to learn a structured world model. \citet{ghugare2022simplifying} argue that aligning the auxiliary objectives and RL objective can be difficult. The authors derive a bound directly on the overall RL objective, which is compatible with self-consistency. The contrastive objectives for model-based RL significantly improve sample efficiency, but they are not only limited to model-based RL. \citet{laskinCURLContrastiveUnsupervised2020} propose replacing the transition function with data augmentations to get a contrastive target and performing model-free learning using the representation produced by the encoder trained with the contrastive target.

@@

\textbf{Reward prediction is not necessary for representation learning}
Unlike prior methods such as TCLR \cite{zhaoSimplifiedTemporalConsistency2023}, TD-MPC \cite{hansenTemporalDifferenceLearning2022}, and Dreamer-V2 \cite{hafnerMasteringAtariDiscrete2022}, our representation learningloss (\cref{eq:rep-loss}) does not include a term for learning to predict the reward or the value. Instead, we rely solely on the self-predictive temporal consistency loss. To analyze the impact of not including the reward prediction term, we perform an ablation study, where we compare our method to three variants of our method:
\begin{itemize}
	\item \ourrew, where we add a reward prediction head into our model architecture by re-defining the dynamics as $\hat{z}_{t+1}, \hat{r}_t = d_{\phi} (z_{t}, a_{t})$ (see \cref{eq:transition}), and include a reward prediction term (discounted MSE Loss) in the representation loss:
	\begin{align*}
		\mathcal{L}_\text{rew} = \sum_{h=0}^{H-1} \gamma^h \| \hat{r}_{t+h} - r_{t+h}\|_2^2,
	\end{align*}
	where $r_{t+h}$ is the ground-truth H-step reward and $\hat{r}_{t+h}$ the predicted H-step rewards.
	\item \ourcosrew, where in addition to the reward prediction head in \ourrew, we additionally convert the self-supervised MSE loss (see \cref{eq:rep-loss}) for the temporal consistency into a cosine similarity, similarly to TCRL:
	\begin{align*} \label{eq:rep-loss}
		\mathcal{L}_{\text{rep}}(\theta, \phi; \tau)
		&= - \sum_{h=0}^{H-1} \gamma^{h} \left ( \frac{\hat{z}_{t+h}}{\| \hat{z}_{t+h} \|_2} \right )^T \left ( \frac{\tilde{z}_{t+h}}{\| \tilde{z}_{t+h} \|_2} \right ).
	\end{align*}
  % where as before we have $\hat{z}_{t+h} = d_{\phi}(\hat{z}_{t}, a_{t})$ and $\tilde{z}_{t+h} = e_{\bar{\theta}}(s_{t+1})$.
	\item TCRL-ours, where in addition to \ourcosrew, we remove the FSQ normalization scheme to make the method equal to TCRL with some minor improvements to the method, such as using nstep-TD3 \cite{fujimotoAddressingFunctionApproximation2018} as the base algorithm instead of DDPG with Double Q and nstep returns, using the neural network architecture from \citet{hansen2023td}, and performing policy smoothing \cite{fujimotoAddressingFunctionApproximation2018}.
\end{itemize}
The results for this ablation study are shown in \cref{fig:reward-ablation}. 
The plots show that our method, \our, without a reward prediction term in the loss, has equal or superior performance to the comparison methods except in Fish Swim, where the unnormalized TCRL-ours outperforms all the methods with normalization due to being able to maintain the latent matrix rank in this specific environment (see \cref{fig:rank-comparison-z1}). Even then, \our performs equally to the normalized variants of our method with the reward prediction. Our results imply that learning to predict the reward is not necessary for learning a suitable latent representation. The upside of not including the reward prediction head is that it makes the representation task-agnostic, which we believe to be important for downstream applications such as learning in an incremental multi-task setting.

@@

\textbf{Reconstruction loss has a detrimental impact}
Learning to minimize the observation reconstruction error has been widely applied in model-based RL \cite{sutton2018reinforcement, haRecurrentWorldModels2018, hafnerLearning2019}, and an observation decoder has been a component of many of the most successful RL algorithms to date \cite{hafner2023mastering}. However, recent work \citep{zhaoSimplifiedTemporalConsistency2023,hansenTemporalDifferenceLearning2022}
has shown that incorporating a reconstruction term into the representation loss can have a negative impact on performance, as learning to reconstruct the observations is inefficient due to the observations containing irrelevant details and visuals like shading that are uncontrollable by the agent and do not affect the tasks.

To provide a thorough analysis of \our, we include results where we add a reconstruction term to our representation loss \cref{eq:rep-loss}:
\begin{align*}
	\mathcal{L}_o &= \mathbb{E}_{o_t\sim\mathcal{D}} [\| \hat{o}_t - o_t \|_2^2],
\end{align*}
where $\hat{o}_t = g_\xi(z_t)$ is a learned observation decoder that takes the latent state as the input and outputs the reconstructed observation. The decoder $g_\xi$ is a standard MLP. We perform reconstruction at each time step in the horizon.
% $g_\xi$ is a learned observation decoder that takes the latent state as the input and outputs the reconstructed observation. The decoder $g_\xi$ is a standard MLP. We perform reconstruction at each time step in the horizon.
The results in \cref{fig:reconstruction-loss-ablation} show that in no environments does reconstruction aid learning, and in some environments, such as the difficult Dog Walk and Quadruped Run, including the reconstruction term has a significant detrimental effect on the performance.
This supports the observations of \citet{zhaoSimplifiedTemporalConsistency2023} and \citet{hansenTemporalDifferenceLearning2022} about the lack of need for a reconstruction target in continuous control tasks.
%Moreover, in the quadruped-run and dog-walk environments incorporating a reconstruction term has a detrimental impact on
%overall performance.
